---
title: "ETC5521 Assignment 1"
subtitle: "Employment and Earnings"
team: Cassowary
author:
  - Kexin Xu
  - Lintian Zhang
date: "`r Sys.Date()`"
output: 
  html_document
---

[This assignment is for ETC5521 Assignment 1 by Team `r rmarkdown::metadata$team` comprising of `r knitr::combine_words(rmarkdown::metadata$author)`.]{style="color:#006DAE;"}

```{r}


employed <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-02-23/employed.csv')

earn <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-02-23/earn.csv')


```




# Introduction and motivation

[FILL] Give the bigger picture of the data, and inspire the reader to learn more about the problem by reading your analysis. 

# Data description

The datasets come from [BLS](https://www.bls.gov/cps/tables.htm#charemp_m), specifically table *cpsaat17* across several years.
The **employed** dataset tells about employed persons by industry, sex, race, and occupation through 2015 to 2020.
```{r}
tribble(
  ~Variable, ~Type,
  "industry", "character",
  "major_occupation", "character",
  "minor_occupation", "character",
  "race_gender", "character",
  "industry_total", "double",
  "employ_n", "double",
  "year", "integer"
)
```

The **earn** dadtaset tells about weekly median earnings and number of persons employed by race/gender/age group through 2010 to 2020.
```{r}
tribble(
  ~Variable, ~Type,
  "sex", "character",
  "race", "character",
  "ethnic_origin", "character",
  "age", "character",
  "year", "integer",
  "quarter", "integer",
  "n_persons", "double",
  "median_weekly_income", "integer"
)
```
The datasets are collected from the Current Population Survey (CPS) which is a monthly survey of households conducted by the Bureau of Census for the Bureau of Labor Statistics.

Based on the datasets, three questions are going to be explored and analyzed in the following section, including:

- what are the changes of people employed in different industries from 2015 to 2020?
 
- what are the demographic differences between industries from 2015 to 2020?

- how do different factors affect the income between 2010 and 2020?

# Analysis and findings

[FILL] Should include at least one plot or numerical summary for each of your questions, that helps the reader arrive at an answer. You should also write paragraphs describing the methods, summaries and findings. 

# References


